A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-09-30 paper/project.tex
Fixing rods-energy dat files and figure, adding conclusion text, other text edits. Tested everything.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the
fundamental-measure theory functional--which describes hard sphere fluids--with
attractive interactions based on the Statistical Associating Fluid Theory
(SAFT). Using this approximation, we are able to describe the vapor pressure, 
vapor density, liquid density, surface tension, enthalpy and
entropy of water, as well as the number of hydrogen bonds and density oscillations
formed near
liquid-vapor and liquid-solid interfaces. These interfaces include water
confined to a hard-walled slit, a single hydrophobic rod immersed in water, and
two hydrophobic rods 
at various separations. For two rods being pulled apart, a transition occurs 
\fixme{at $(\pi-2)r$?} from a vapor state to a liquid state in the area between
the rods. Even for rods of a small radius (0.3~-~0.9~nm), we find that for the
vapor state, the force can be approximated
classically and is equal to 144~mN/m, or twice the bulk surface tension.
This functional is computationally efficient and provides a foundation for 
future work describing interactions in aqueous solution on an
atomic level.}

@@

The dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp} &= A_1 + \beta A_2
\end{align}
where $A_1$ and $A_2$ \fixme{change all to $a_1$ and $a_2$} are the first 
two terms in a high-temperature
perturbation expansion and $\beta=\frac{1}{k_BT}$.  $A_1$ is the mean-field dispersion
interaction, which reduces in the homogeneous limit to
\begin{align}\label{eq:A1-simple}
  \frac{A_1}{V} &= \frac12 n_b \int \varphi(\left|\xx\right|)
  g_{HS}(\left|\xx\right|) d\xx
\end{align}
where $n_b$ is the bulk density and $g_{HS}$ is the homogeneous
hard-sphere fluid correlation function.
The second dispersion term in the free energy $A_2$ describes the
effect of fluctuations resulting from the compression of the fluid due
to the dispersion interaction itself, and is commonly approximated
using the local compressibility approximation (LCA). In the LCA,
we assume the energy fluctuation is directly related to the
compressibility of a hard-sphere reference fluid\cite{barker1976liquid}.

@@

The form of $A_1$ is given in
reference~\cite{gil-villegas-1997-SAFT-VR}, but is expressed in terms
of the filling fraction.  In order to apply this form to an
inhomogeneous density distribution, we construct an effective local
filling fraction for dispersion $\etadisp$, given by
\begin{align}
  \etadisp(\xx) &=
  \int \frac{\Theta(2\lambdadisp\lscale R - \left|\xx-\xx'\right|)}{
             \left(2\lambdadisp\lscale\right)^3} n(\xx') d\xx'.
\end{align}
This effective filling fraction is used throughout the dispersion
calculations, and represents a filling fraction averaged over the
effective range of the dispersive interaction.  This functional
introduces an additional empirical parameter $\lscale$ which adjusts
the length scale over which the dispersion interaction is correlated.

The first term in the dispersion functional $A_1$ when written using
the above filling fraction for dispersion has the form
\begin{align}
  A_1 &= \int
   -4(\lambdadisp^3 - 1)\epsilondisp \etadisp(\xx)
    g^{HS}_\sigma(\eta_\textit{eff}(\xx))
  d\xx \\
  \eta_\textit{eff}(\xx) &= \sum_{ij=0}^3 C_{ij}  \etadisp^i(\xx)
  \lambdadisp^j
  \\
  g_\sigma^{HS}(\eta) &= \frac{1}{1-\eta}
  +\frac32\frac{\zeta(\xx)}{(1-\eta)^2}
  + \frac12\frac{\eta^2\zeta(\xx)}{(1-\eta)^3}
  \\
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}
\end{align}
where $C_{ij}$ are numerical constants taken from
reference\cite{gil-villegas-1997-SAFT-VR}, which come from a numerical
fit to the integral in Equation~\ref{eq:A1-simple} over a range of
values for $\lambdadisp$ and filling fractions.  Here $g_\sigma^{HS}$
is the hard-sphere fluid correlation function evaluated at contact and uses the
effective filling fraction for inhomogeneous density distributions.
The $\zeta$ factor
(also seen in the association energy, below) is a dimensionless
measure of the local density inhomogeneity.

The second term, $A_2$, which describes the contribution to the free
energy associated with fluctuations is given by
\begin{align}
  A_2 &= \int \frac12 \epsilondisp
              K_{HS}(\etadisp(\xx))
              \etadisp(\xx)
              \frac{\delta A_1}{\delta \etadisp(\xx)}
                d\xx
\end{align}
where $K_{HS}$ is the Percus-Yevick hard-sphere compressibility, given
by
\begin{align}
  K_{HS} &=
    \frac{(1 - \etadisp)^4}{1 + 4(\etadisp + \etadisp^2)}.
\end{align}

@@

The density profile for a slit of size 4.0~nm is shown in Figure~\ref{fig:energy-1D}. 
Near the interface, there are oscillations around the
liquid density which decrease near the center of the slit. We expect the center
of the slit to act approximately like bulk water, and the absence of
oscillations supports this. For larger slits (6.0~nm or more) the majority of
the water molecules in the slit act like the bulk. Based on our computations, we
also determine the smallest slit that can contain liquid water. This result is
approximately 0.6~nm. For smaller slits, the initial liquid water inside will
vaporize. The results for larger liquid-filled slits are only metastable,
however, as a vapor-filled slit is thermodynamically favorable. Based on our
calculations for a pressure of 1~atm, the work needed to pull apart
the two hard walls comprising the slit is one to two orders of magnitude less
than the free energy of the liquid-filled slit. For the range of slit sizes we
studied, all the computational results of liquid-filled slits are
metastable. Lum, Chandler and Weeks \cite{lum1999hydrophobicity} discuss the
stability of water between large hydrophobic plates and the critical separation
necessary for stability. For water at room temperature and atmospheric pressure,
plates with a separation smaller than about 100~nm will not contain stable liquid
water. They go on to say that confined water has a region of metastability down
to separations of about 5~nm. This is comparable to the slit separation where
our calculations show water inside the slit acting like the bulk liquid, which
we know to be a metastable solution. Forsman \emph{et al.}\cite{forsman1996computer}
also study water between two hard walls at constant chemical potential. They 
also conclude that the liquid state is likely metastable, and there is a free
energy barrier to reach the stable vapor phase.

@@

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of various radii at separation $d$ (see Figure~\ref{fig:rods}), ranging from $d=0$ 
(touching rods) to $d=1$~nm. At small
separations there is only vapor between the rods. As the rods are
pulled apart, the vapor region expands until some critical separation is
reached and liquid water fills the region between the rods. Figure~\ref{fig:density-rods} 
shows density profiles before and after this transition
for rods of radius 0.5~nm. This critical separation for the transition to liquid depends
on the radii of the rods, and is about 0.55~nm for the rods shown in 
Figure~\ref{fig:density-rods}.
\fixme{add analysis of circumferences of stadium shape vs two rods and
calculating the expected critical separation}
Since we are looking at only hard-walled
non-interacting rods, the critical separation will be
different for any system where there is an appreciable force between rods. 
The density oscillations seen around the rods in Figure~\ref{fig:density-rods} 
are very similar to
those for the single hydrophobic rod in water (see Figure~\ref{fig:density-single-rod}).
The oscillations around the two rods at small separation looks as if the two rods were one
solid stadium shaped object (approximately like a rectangle with semi-circles on the ends).

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two hydrophobic rods ranging in radius from
0.3~nm to 0.9~nm. \fixme{r=0.9 to be added when it finishes!}
All were arbitrarily set to go to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-rods}. }
\label{fig:rods-energy-vs-distance}
\end{figure}

The free energy per length is compared for rods of several different radii
in Figure~\ref{fig:rods-energy-vs-distance}. In this figure, the energy for a radius 0.5~nm
is the same as the free energy shown in Figure~\ref{fig:energy-rods}.
For comparison, we arbitrarily adjust each data set such that the
free energy per length goes to zero at large separations. At large separations, the rods will 
simply act like two separate single rods in water and the total energy will
reach an asymptote proportional to the surface tension (similar to that shown in 
Figure~\ref{fig:energy-vs-diameter}).
We see that all the rods studied have the same behavior, with a positive linear relation 
between free energy
per length and separation during the initial state where the rods have vapor between them. 
The slope of this line is equal to the force per length on the rods, which we calculate to 
be 144~mN/m, or twice the surface tension
found by the functional (see Figure~\ref{fig:surface-tension}). This indicates that
even though the rods are nanoscale, to the water molecules they can be treated classically
\fixme{reference above equations/explanation with circumference, critical separation, etc.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

\needsworknow{Conclusions about slit, single rod, two hydrophobic rods (mostly). 
computationally efficient. future uses}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the
fundamental-measure theory functional--which describes hard sphere fluids--with
attractive interactions based on the Statistical Associating Fluid Theory
(SAFT). Using this approximation, we are able to describe the vapor pressure, 
saturated liquid density, surface tension, enthalpy and
entropy of water, as well as the number of hydrogen bonds and density oscillations
formed near
liquid-vapor and liquid-solid interfaces. These interfaces include water
confined to a hard-walled slit, a single hydrophobic rod immersed in water, and
two hydrophobic rods 
at various separations. For two rods being pulled apart, a transition occurs 
from a vapor state to a liquid state in the area between
the rods at approximately $(\pi-2)$ multipled by the radius. Even 
for these nanoscale rods (0.3~-~1.0~nm), we find that in the
vapor state, the force can be approximated
classically and is equal to 144~mN/m, or twice the bulk surface tension.
This functional is computationally efficient and provides a foundation for 
future work describing interactions in aqueous solution on an
atomic level.}

@@

The dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp} &= A_1 + \beta A_2
\end{align}
where $a_1$ and $a_2$ \fixme{change all to $a_1$ and $a_2$} are the first 
two terms in a high-temperature
perturbation expansion and $\beta=1/k_BT$.  The first term $a_1$ is 
the mean-field dispersion
interaction, which reduces in the homogeneous limit to
\begin{align}\label{eq:A1-simple}
  \frac{A_1}{V} &= \frac12 n_b \int \varphi(\left|\xx\right|)
  g_{HS}(\left|\xx\right|) d\xx
\end{align}
where $n_b$ is the bulk density and $g_{HS}$ is the homogeneous
hard-sphere fluid correlation function.
The second dispersion term in the free energy $a_2$ describes the
effect of fluctuations resulting from the compression of the fluid due
to the dispersion interaction itself, and is commonly approximated
using the local compressibility approximation (LCA). In the LCA,
we assume the energy fluctuation is directly related to the
compressibility of a hard-sphere reference fluid\cite{barker1976liquid}.

@@

The form of $a_1$ is given in
reference~\cite{gil-villegas-1997-SAFT-VR}, but is expressed in terms
of the filling fraction.  In order to apply this form to an
inhomogeneous density distribution, we construct an effective local
filling fraction for dispersion $\etadisp$, given by
\begin{align}
  \etadisp(\xx) &=
  \int \frac{\Theta(2\lambdadisp\lscale R - \left|\xx-\xx'\right|)}{
             \left(2\lambdadisp\lscale\right)^3} n(\xx') d\xx'.
\end{align}
This effective filling fraction is used throughout the dispersion
calculations, and represents a filling fraction averaged over the
effective range of the dispersive interaction.  This functional
introduces an additional empirical parameter $\lscale$ which adjusts
the length scale over which the dispersion interaction is correlated.

The first term in the dispersion functional $a_1$ when written using
the above filling fraction for dispersion has the form
\begin{align}
  A_1 &= \int
   -4(\lambdadisp^3 - 1)\epsilondisp \etadisp(\xx)
    g^{HS}_\sigma(\eta_\textit{eff}(\xx))
  d\xx \\
  \eta_\textit{eff}(\xx) &= \sum_{ij=0}^3 C_{ij}  \etadisp^i(\xx)
  \lambdadisp^j
  \\
  g_\sigma^{HS}(\eta) &= \frac{1}{1-\eta}
  +\frac32\frac{\zeta(\xx)}{(1-\eta)^2}
  + \frac12\frac{\eta^2\zeta(\xx)}{(1-\eta)^3}
  \\
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}
\end{align}
where $C_{ij}$ are numerical constants taken from
reference\cite{gil-villegas-1997-SAFT-VR}, which come from a numerical
fit to the integral in Equation~\ref{eq:A1-simple} over a range of
values for $\lambdadisp$ and filling fractions.  Here $g_\sigma^{HS}$
is the hard-sphere fluid correlation function evaluated at contact and uses the
effective filling fraction for inhomogeneous density distributions.
The $\zeta$ factor
(also seen in the association energy, below) is a dimensionless
measure of the local density inhomogeneity.

The second term, $a_2$, which describes the contribution to the free
energy associated with fluctuations is given by
\begin{align}
  A_2 &= \int \frac12 \epsilondisp
              K_{HS}(\etadisp(\xx))
              \etadisp(\xx)
              \frac{\delta A_1}{\delta \etadisp(\xx)}
                d\xx
\end{align}
where $K_{HS}$ is the Percus-Yevick hard-sphere compressibility, given
by
\begin{align}
  K_{HS} &=
    \frac{(1 - \etadisp)^4}{1 + 4(\etadisp + \etadisp^2)}.
\end{align}

@@

The density profile for a slit of size 4.0~nm is shown in Figure~\ref{fig:energy-1D}. 
Near the interface, there are oscillations around the
liquid density which decrease near the center of the slit. We expect the center
of the slit to act approximately like bulk water, and the absence of
oscillations supports this. For larger slits (6.0~nm or more) the majority of
the water molecules in the slit act like the bulk. \fixme{Either add brief
description of bottom half of Figure~\ref{fig:energy-1D} or delete it}
Based on our computations, we
also determine the smallest slit that can contain liquid water. This result is
approximately 0.6~nm. For smaller slits, the initial liquid water inside will
vaporize. The results for larger liquid-filled slits are only metastable,
however, as a vapor-filled slit is thermodynamically favorable. Based on our
calculations for a pressure of 1~atm, the work needed to pull apart
the two hard walls comprising the slit is one to two orders of magnitude less
than the free energy of the liquid-filled slit. For the range of slit sizes we
studied, all the computational results of liquid-filled slits are
metastable. Lum, Chandler and Weeks \cite{lum1999hydrophobicity} discuss the
stability of water between large hydrophobic plates and the critical separation
necessary for stability. For water at room temperature and atmospheric pressure,
plates with a separation smaller than about 100~nm will not contain stable liquid
water. They go on to say that confined water has a region of metastability down
to separations of about 5~nm. This is comparable to the slit separation where
our calculations show water inside the slit acting like the bulk liquid, which
we know to be a metastable solution. Forsman \emph{et al.}\cite{forsman1996computer}
also study water between two hard walls at constant chemical potential. They 
also conclude that the liquid state is likely metastable, and there is a free
energy barrier to reach the stable vapor phase.

@@

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of radii 0.3~-~1.0~nm at separation $d$ (see Figure~\ref{fig:rods}), ranging from $d=0$ 
(touching rods) to $d=1$~nm. At small
separations there is only vapor between the rods. As the rods are
pulled apart, the vapor region expands until some critical separation is
reached and liquid water fills the region between the rods. Figure~\ref{fig:density-rods} 
shows density profiles before and after this transition
for rods of radius 0.5~nm. This critical separation for the transition to liquid depends
on the radii of the rods, and is about 0.55~nm for the rods shown in 
Figure~\ref{fig:density-rods}. 
Since we are looking at only hard-walled
non-interacting rods, the critical separation will be
different for any system where there is an appreciable force between rods. 
The density oscillations seen around the rods in Figure~\ref{fig:density-rods} 
are very similar to
those for the single hydrophobic rod in water (see Figure~\ref{fig:density-single-rod}).
The oscillations around the two rods at small separation looks as if the two rods were one
solid stadium shaped object (approximately like a rectangle with semi-circles on the ends).

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two hydrophobic rods ranging in radius from
0.3~nm to 1.0~nm.
All were arbitrarily set to go to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-rods}. }
\label{fig:rods-energy-vs-distance}
\end{figure}

The free energy per length is compared for rods of several different radii
in Figure~\ref{fig:rods-energy-vs-distance}. In this figure, the energy for a radius 0.5~nm
is the same as the free energy shown in Figure~\ref{fig:energy-rods}.
For comparison, we arbitrarily adjust each data set such that the
free energy per length goes to zero at large separations. At large separations, the rods will 
simply act like two separate single rods in water and the total energy will
reach an asymptote proportional to the surface tension (similar to that shown in 
Figure~\ref{fig:energy-vs-diameter}).
We see that all the rods studied have the same behavior, with a positive linear relation 
between free energy
per length and separation during the initial state where the rods have vapor between them. 
The slope of this line is equal to the force per length on the rods, which we calculate to 
be 144~mN/m, or twice the surface tension
found by the functional (see Figure~\ref{fig:surface-tension}). This indicates that
even though the rods are nanoscale, to the water molecules they can be seen classically.
To estimate how the critical transition depends on the radius,
we assume a classical approximation where the energy on the rod is equal to the
surface area multiplied by the bulk surface tension. The surface area is just the
circumference multiplied by the width, and this width has the same value for both the
vapor-filled stadium shape and the liquid-filled separate rods. We propose that 
the critical separation
for the transition, $d_c$, will occur when the circumference of the stadium shape (see 
top of Figure~\ref{fig:density-rods}) and the circumference of the two rods (see bottom
of Figure~\ref{fig:density-rods}) are equal:
\begin{align}
C_{stadium} &= C_{rods} \\
2\pi r+4r+2d_c &= 4\pi r \\
d_c &= (\pi-2)r.\label{criticalseparation}
\end{align}

\begin{table}
\begin{tabular} {|c|c|c|c|}
\hline
$r$ (nm) & $d_c$ prediction (nm) & $d_c$ result (nm) & \% error \\
\hline
0.3 & 0.34 & 0.25 & 26\% \\
\hline
0.5 & 0.57 & 0.55 & 3.5\% \\
\hline
0.7 & 0.80 & 0.83 & 3.8\% \\
\hline
0.9 & 1.02 & 1.07 & 4.9\% \\
\hline
1.0 & 1.14 & 1.18 & 3.5\% \\
\hline
\end{tabular}
\caption{Calculations of the critical separation (see Equation~\ref{criticalseparation})
for the transition
from liquid to vapor between two hydrophobic rods.}
\label{criticalseparationtable}
\end{table}
The predicted and actual computed values for $d_c$ are shown in 
Table~\ref{criticalseparationtable},
and they match fairly well for the 0.5~-~1.0~nm rods. We hypothesize that
larger radii rods
would also follow this trend, but smaller rods are likely to have a different
relationship with the surrounding water molecules. This may explain why the critical
separation result for a radius of 0.3~nm does not match the prediction well.

@@

\section{Conclusion}

The classical density functional we developed by combining fundamental-measure 
theory with SAFT is able to produce accurate results for hydrophobic interations.
\fixme{decribe any results for slit?}
We added the parameter \lscale to the functional and fit it to the surface tension
at room temperature, which we find to be consistent with results for hydrophobic
rods. The free energy per area for water surrounding a hydrophobic rod shows the
expected behavior with an asymptote equal to the surface tension.
Based on this result, we expect to be
able to describe classically rods with a radius of about 0.5~nm or larger. Even
though this is still nanoscale, from the perspective of the water molecules
the interaction with a hydrophobic rod is easily approximated classically.
For the interactions between water and two hydrophobic rods, we see that a
classical treatment of the phase transition critical separation works well
for rods larger than 0.5~nm, within 5\%. This transition from a vapor-filled
stadium shape to two rods with liquid between them (see Figure~\ref{fig:density-rods})
is commonly seen for objects of a similar geometry. Further evidence for the
accuracy of this classical treatment is the calculation of the force on the
stadium shape. The force per length is constant and equal to twice the surface tension.

This functional produces many of the results of \emph{ab initio} calculations
while being much more computationally efficient. This allows for future work
involving hydrophilic interactions with more complex electrostatic potentials.
\fixme{more about efficiency or future work?}
